Type Visualization for R

### 2020-06-21 paper/main.tex
paper: My edits to abstract and intro

\abstract{
  Data-driven approaches to programming language design are uncommon.
  Despite the availability of large code repositories, distilling useful
  information from programs remain difficult. Important dimensions of
  code, like run-time type data, are inscrutable without appropriate tools.
  We built \typevis to explore and analyze run-time types in
  the R programming language by means of visualization.
  Our system focuses on understanding and comparing the type signatures of
  function calls across the R ecosystem.
  \typevis uses flows to intuitively represent categorical
  and numerical aspects of type signatures.
  Insights derived from our visualization are aimed at informing
  language design decisions---specifically of a gradual type system
  currently being developed for R.
}

@@

Programming languages commonly evolve by decree. Often, the designer
decides that a new feature is necessary or that a past feature was
ill-conceived. Thus, the language moves forward,
forcing its users to adapt to the changes.
Sometimes, these changes are the result of rigorous theory
or the product of community feedback. However, rarely is language design
informed by empirical data about how programs are \emph{actually
written} in practice.

%For a popular language, data collection is no challenge. Vast
Thanks to the large amount of open source code, collecting data about
the usage of a popular programming language is feasible. Vast
quantities of publicly available code are hosted on open source
repositories such as Github.
Furthermore, if there is a concern about the quality of such data
(for example, code duplication~\cite{lopes:2017}),
code can be obtained from curated
language-specific package repositories such as npm or CRAN.

To inform programming language design, all the collected data
needs to be analyzed and interpreted.
As code is complex and highly structured, simple semantically-ignorant
metrics like ``lines of code'' carry little significance.
Therefore, programming languages researchers often employ static
and dynamic analyses to gather information about specific aspects
of programs in the dataset, for example, the use of a language feature.
And even then it may be hard to make sense of the data,
especially if the dataset is large.
%Rather, the difficulty lies with analyzing and interpreting the data
%at hand. Code is complex and highly structured. Simple

To analyze one aspect of programming languages---\emph{the use of types}---we
built \typevis, an interactive visualization of run-time type
signatures of function calls.
%an interactive visualization of run-time types for the language R.
\typevis is instantiated for the language R:
it builds on a dataset of concrete execution traces over unit tests and vignettes
of the most widely used libraries in the R ecosystem.
For example, for the function call \code{!(FALSE)},
the recorded type signature would be \code{logical->logical},
meaning that function \code{!} took a logical value and returned a logical value.

The visualization supports three broad tasks:
\begin{itemize}
  \item filtering the dataset down to an amenable subset;
  \item understanding the types of function arguments and returns;
  \item comparing types across different functions.
\end{itemize}
%For filtering, \typevis uses the treemap idiom
%to convey the relative scale of different packages.
For exploring and filtering the dataset, \typevis relies on the treemap idiom
to represent R packages and functions, with the block sizes conveying
the relative popularity in the dataset.
In addition to the treemaps, a search box can be used to directly filter
by functions.
Once a specific subset of one or more functions is selected,
their type signatures are represented as flows over types
where the width of the flow is proportional to frequency.
These flows can be readily compared within the same function,
or across different functions.

Tasks enabled by \typevis can assist language designers
during multiple phases of development.
For example, exploratory data analysis can identify unexpected uses
of specific functions, or weed out language designs
that are incompatible with existing programs.
In particular, the visualization is built to answer questions
relevant to the development of a gradual type system for R.
However, the visualization is not limited to R and can be used to
analyze type usage in any programming language where type information
is available at run-time.
Supplemental material, including the data and source code of
\typevis, is freely available: \todo{Fill with archival link}

@@

Each execution trace was recorded by a dynamic analysis
running the test, example, and vignette code of each package.
A vignette is a form of documentation that weaves
together prose and executable code.
An execution trace is a list of ``raw'' function call signatures
containing, among other attributes, the function name
and type tags for each argument and return value.
The type tags are assigned by the dynamic analysis and
refine over actual type tags in R; for example, the analysis
assigns scalar tag \code{double} to a singleton vector of double values
because R does not have native support for scalars.
Our dataset consists of a table of reduced execution traces:
because \typevis is designed for the analysis of types, not all information
in the traces is relevant.
%that have been processed to reduce some fields.
%Traces were generated via a dynamic analysis
%that instruments the R virtual machine with hooks
%that intercept the required data at run-time.

@@

\typevis is an interactive web-based visualization
that permits real-time exploration,
even with large datasets.
The default dataset contains over a million rows
based on the analysis of almost five hundred
different R packages.
Figure~\ref{fig:typevis} displays
the heart of \typevis
in its initial configuration.
Most aspects of the system can be tweaked
by a user if desired.

@@

Treemaps are often used for displaying hierarchical data~\cite{shneiderman:1992}.
Package systems in many languages,
like Java,
contain deeply nested namespace hierarchies.
This idiom is appropriate for navigation of such
hierarchies.
R packages are not structured deeply
so the treemaps in our data are only one
level deep.
However, if the dataset was more deeply nested,
\typevis can be configured to
handle that.

@@

We have presented \typevis, an interactive visualization
of run-time type information for R.
It supports real-time exploration of a dataset of
execution traces from a large corpus of R packages.
To narrow the data under consideration,
\typevis provides facilities for filtering
and searching.
Type signatures are represented as flows traversing across
each argument type and return type in sequence.
Flow widths are proportional to how often
that type signature was seen during the dynamic analysis.
Interactions provide additional details and
assist users in comprehension and navigation.

Our tool is aimed at elucidating how R programmers
use the language in practice.
Specifically, the data and visualization will be used to
inform the design of a gradual type system for R.
However, \typevis is not tied to the R ecosystem
and is suitable to use for any dynamically typed language
with a similar dataset.
Additionally, the visualization could yield insights
useful for other purposes than gradual typing.

>>>>

\abstract{
  Data-driven approaches to programming language design are uncommon.
  Despite the availability of large code repositories, distilling useful
  information from programs remains difficult. Important dimensions,
  like run-time type data, are inscrutable without the appropriate tools.
  \typevis is an interactive visualization for exploring and analyzing
  type information from program execution traces---aiding
  user understanding of function type signatures across many executions.
  The system uses flows to intuitively represent categorical
  and numerical aspects of type signatures.
  Insights derived from our visualization are aimed at informing
  language design decisions---specifically of a new gradual type system
  currently being developed for the R programming language.
}

@@

Programming languages commonly evolve by decree. Often, the language designer
decides that a new feature is necessary, or that a past feature was
ill-conceived. Thus, the language moves forward---
forcing its users to adapt to the changes.
Sometimes, these changes are the result of rigorous theory
or the product of community feedback. However, rarely is language design
informed by empirical data on how programs are \emph{actually
written} in practice. Why so?

%For a popular language, data collection is no challenge. Vast
Thanks to the prevalence of open source code, collecting data on
the usage of a popular programming language is feasible. Vast
quantities of publicly available code are hosted on repositories such as Github.
Furthermore, if there is concern about the quality of this data,
such as code duplication~\cite{lopes:2017},
code can be obtained from curated
language-specific package repositories like npm or CRAN.

To inform programming language design,
the collected data needs to be analyzed and interpreted.
Programs are complex and highly structured,
so simple semantically-ignorant metrics like ``lines of code'' carry little significance.
Therefore, researchers often employ static and dynamic analyses
to gather information about specific aspects of programs.
Even then, it may be difficult to make sense of the
results of these analyses,
especially if the data set is large.
%Rather, the difficulty lies with analyzing and interpreting the data
%at hand. Code is complex and highly structured. Simple

To analyze just one aspect of programming languages, \emph{type usage}, we
built \typevis, an interactive visualization of run-time type signatures.
%an interactive visualization of run-time types for the language R.
\typevis was created with the R programming language in mind,
and builds on a data set of execution traces
mined from the unit tests and vignettes
of the most widely used libraries in the R ecosystem.

The tool supports three broad tasks:
\begin{itemize}
  \setlength\itemsep{0em}
  \item filtering the data set down to an amenable subset;
  \item understanding the types of function arguments and returns;
  \item comparing type signatures across different functions.
\end{itemize}
%For filtering, \typevis uses the treemap idiom
%to convey the relative scale of different packages.
For exploring and filtering the data set,
\typevis relies on treemaps to represent R packages and functions,
with node size conveying relative popularity.
Additionaly, one can use a search box to directly filter
by function name.
Once a subset of one or more functions is selected,
their type signatures are encoded as flows over types,
where flow width is proportional to frequency.
These flows can be readily compared within the same function,
or between different functions.

\typevis can assist language designers during multiple phases of development.
For example, exploratory data analysis can identify unexpected edge cases,
or weed out language designs that are incompatible with existing programs.
\typevis was built in support of a new gradual type system for R,
whose adoption depends on integrating well with existing R code.
However, the visualization is not limited to R and can be used to
analyze type usage in any language where similar data is available.

Supplemental material, including the data and source code of
\typevis, is freely available: \todo{Fill with archival link}

@@

Each execution trace was recorded by a dynamic analysis
running the test, example, and vignette code of each package.
A vignette is a form of documentation that weaves
together prose and executable code.
An execution trace is a list of ``raw'' function call signatures
containing, among other attributes, the function name
and type tags for each argument and return value.
For example, for the function call \code{!(FALSE)},
the recorded type signature would be \code{logical->logical},
meaning that function \code{!} took a logical value and returned a logical value.
The type tags are assigned by the dynamic analysis and
refine over actual type tags in R; for example, the analysis
assigns scalar tag \code{double} to a singleton vector of double values
because R does not have native support for scalars.
Our data set consists of a table of reduced execution traces:
because \typevis is designed for the analysis of types, not all information
in the traces is relevant.
%that have been processed to reduce some fields.
%Traces were generated via a dynamic analysis
%that instruments the R virtual machine with hooks
%that intercept the required data at run-time.

@@

\typevis is an interactive web-based visualization
that permits real-time exploration,
even with large data sets.
The default data set contains over a million rows
based on the analysis of almost five hundred
different R packages.
Figure~\ref{fig:typevis} displays
the heart of \typevis
in its initial configuration.
Most aspects of the system can be tweaked
by a user if desired.

@@

Treemaps are often used for displaying hierarchical data~\cite{shneiderman:1992}.
Package systems in many languages,
like Java,
contain deeply nested namespace hierarchies.
This idiom is appropriate for navigation of such
hierarchies.
R packages are not structured deeply
so the treemaps in our data are only one
level deep.
However, if the data set was more deeply nested,
\typevis can be configured to
handle that.

@@

We have presented \typevis, an interactive visualization
of run-time type information for R.
It supports real-time exploration of a data set of
execution traces from a large corpus of R packages.
To narrow the data under consideration,
\typevis provides facilities for filtering
and searching.
Type signatures are represented as flows traversing across
each argument type and return type in sequence.
Flow widths are proportional to how often
that type signature was seen during the dynamic analysis.
Interactions provide additional details and
assist users in comprehension and navigation.

Our tool is aimed at elucidating how R programmers
use the language in practice.
Specifically, the data and visualization will be used to
inform the design of a gradual type system for R.
However, \typevis is not tied to the R ecosystem
and is suitable to use for any dynamically typed language
with a similar data set.
Additionally, the visualization could yield insights
useful for other purposes than gradual typing.
